% Copyright 2026 the author. All rights reserved.

\documentclass{article}
\usepackage[letterpaper, textwidth=5.0in, textheight=9.0in]{geometry}

\frenchspacing

\usepackage{amsmath, amsthm, amsfonts}
\let\Re\relax 
\DeclareMathOperator{\Re}{Re}
\DeclareMathOperator{\dd}{d\!}
\newcommand{\mean}[1]{\langle{#1}\rangle}
\newcommand{\given}{\,|\,}
\theoremstyle{definition}
\newcommand{\problemname}{Problem}
\newtheorem{problem}{\problemname}
\newcommand{\probref}[1]{\problemname~\ref{#1}}
\newtheorem{conjecture}{Conjecture}[problem]
\newtheorem{note}{Note}[problem]

\begin{document}
    
\section*{Group averaging of exoplanet likelihoods}
\noindent
\textit{by} \textbf{Hogg} \textit{for} \textbf{Shu} \textit{with help from} \textbf{Villar}.

\paragraph{Abstract:}
The idea is to figure out how to do group averages (or Bayesian mar\-gin\-a\-li\-za\-tions) of likelihood functions, very fast.

\begin{problem}[absolute toy]\label{prob:toy}
Imagine that we have a code handle to a scalar function $f(\phi)\in\mathbb{R}$ that is a function of an angle $\phi\in U(1)$.
We don't know the form or value of this function \textsl{a priori} but we have code that can evaluate the function at any angle $0\leq\phi<2\pi$ or at any list of angles.
We believe that this function could be reasonably well approximated by a low-degree polynomial of trig functions, or
\begin{equation}
    f(\phi) \approx \Re\left\{\sum_{m=0}^M a_m\,\exp(i\,m\,\phi)\right\} ~,
\end{equation}
where $M$ is a small integer, and the $a_m$ are complex numbers (which we don't know).
We seek to find an estimate or value of the average $\mean{f}$ or marginalization
\begin{equation}
    \mean{f} = \frac{1}{2\pi}\,\int_0^{2\pi} f(\phi)\,\dd\phi
\end{equation}
taking as few evaluations of $f(\phi)$ as possible.
Additionally, we would like some assessment of the quality of the approximation, or at least a flag if the approximation is poor.
\end{problem}

\begin{conjecture}
If $f(\phi)$ is indeed well approximated by a polynomial of degree $M$, then we can get a good estimate of the average $\mean{f}$ with $2\,M+1$ function evaluations.
\end{conjecture}

\begin{conjecture}
If you want an estimate of $\mean{f}$ and \emph{also} some kind of confidence or check of the integral, you will have to do substantially more than $2\,M+1$ function evaluations.
\end{conjecture}

\begin{note}
There might be a concept of a ``torus design'' that is relevant here.
It is a finite set of points that can be used to exactly average a finite-degree polynomial function of angle.
\end{note}

\begin{problem}[exponentiated toy]\label{prob:exp}
Same as \probref{prob:toy}---we want to know $\mean{f}$---but now it is the logarithm of the function that is well described by the polynomial; that is
\begin{equation}
    \ln f(\phi) \approx \Re\left\{\sum_{m=0}^M a_m\,\exp(i\,m\,\phi)\right\} ~.
\end{equation}
\end{problem}

\begin{conjecture}
The solution will also only take $2\,M+1$ function evaluations, but, subsequently, a numerical integration will be required.
That is, there won't be any simple form for the integral, given the function evaluations.
\end{conjecture}

\begin{conjecture}
If we want to develop a reliable bound on the error on our estimate of the average, we might have to take additional samples near the peak of the function.
Of course thie problem as currently posed doesn't ask for a bound on the error; probably it should.
\end{conjecture}

\begin{problem}[2D toy likelihood function]\label{prob:toylike}
Choose a random (latent) non-zero 2-vector $z\in\mathbb{R}^2$.
Create some ``data'' $y$ to be a noisy estimate of this 2-vector, such that
\begin{equation}
    y = z + \epsilon ~,
\end{equation}
and $\epsilon$ is drawn from a 2-dimensional normal with zero mean and variance tensor $V=\sigma^2\,I_2$, where $I_2$ is the $2\times 2$ identity matrix.
Now imagine that we are trying to estimate $z$, given the data $y$, and knowing $\sigma$.
Write code to compute the likelihood function $p(y\given z)$.

Since we can parameterize the vector $z$ in terms of a magnitude $r$ and an angle $\phi$ (with respect to a coordinate axis, say), this likelihood function can be written as $p(y\given r,\phi)$, which is a function of $\phi$.
Find the average of this function using the techniques developed in \probref{prob:toy} and \probref{prob:exp}.
What issues do you encounter and how do you find $M$ values that are reasonable?
How should you plot or present your results? Treat the data $y$ to be fixed, but the magnitude $r$ to be variable.
\end{problem}

\begin{conjecture}
The value of $M$ will probably depend on whether you are using your solution to \probref{prob:toy} or to \probref{prob:exp}.
It will also depend on the ratio $|z|/\sigma$ of the magnitude $|z|$ of the true $z$ you chose at the beginning to the uncertainty root-variance $\sigma$.
This quantity is what an astronomer might call ``the signal-to-noise ratio.''
\end{conjecture}

\begin{problem}[3D toy likelihood function]\label{prob:toylike3}
Generalize \probref{prob:toylike} to be in $\mathbb{R}^3$.
Now instead of one angle $\phi$, there are two angles $\theta,\phi$ on the sphere.
The polynomial now will not be in the basis of $\exp(i\,m\,\phi)$ but instead in terms of spherical harmonics $Y_{\ell m}(\theta,\phi)$.
The complexity parameter will now be $L$, the maximum degree $\ell$ of the spherical harmonics.
\end{problem}

\begin{conjecture}
If the likelihood is approximated as a spherical-harmonic polynomial of degree $L$, the solution will require $L^2$ function evaluations.
\end{conjecture}

\begin{note}
It might be worth looking up (on \textit{Wikipedia} or equivalent) the mathematical concept of \emph{spherical designs}.
There might be known methods here that are very good, or else this mathematical literature might be useless.
\end{note}

\end{document}
